# A Primer in Regression Techniques {#regression}

> All models are wrong, but some are useful -- Box

## Introduction

This chapter will provide all the foundations we need for the coming chapters.
It is not intended as a general and all-exhaustive introduction to
regression techniques, but rather the minimum requirement moving forwards.
We will also hone our data processing and plotting skills.

## Prerequisites

```{r regr-libs,message=TRUE,warning=FALSE}
library(mefa4)                # data manipulation
library(mgcv)                 # GAMs
library(pscl)                 # zero-inflated models
library(lme4)                 # GLMMs
library(MASS)                 # Negative Binomial GLM
library(partykit)             # regression trees
library(intrval)              # interval magic
library(opticut)              # optimal partitioning
library(visreg)               # regression visualization
library(ResourceSelection)    # marginal effects
library(MuMIn)                # multi-model inference
source("functions.R")         # some useful stuff
load("_data/josm/josm.rda") # JOSM data
```

Let's pick a species, Ovenbird (`OVEN`), that is quite common and abundant in the data set.
We put together a little data set to work with:

```{r regr-data}
spp <- "OVEN"

ytot <- Xtab(~ SiteID + SpeciesID, josm$counts[josm$counts$DetectType1 != "V",])
ytot <- ytot[,colSums(ytot > 0) > 0]
x <- data.frame(
  josm$surveys, 
  y=as.numeric(ytot[rownames(josm$surveys), spp]))
x$FOR <- x$Decid + x$Conif+ x$ConifWet # forest
x$AHF <- x$Agr + x$UrbInd + x$Roads # 'alienating' human footprint
x$WET <- x$OpenWet + x$ConifWet + x$Water # wet + water
cn <- c("Open", "Water", "Agr", "UrbInd", "SoftLin", "Roads", "Decid", 
  "OpenWet", "Conif", "ConifWet")
x$HAB <- droplevels(find_max(x[,cn])$index) # drop empty levels
x$DEC <- ifelse(x$HAB == "Decid", 1, 0)

table(x$y)
```

## Poisson null model

The null model states that the expected values of the count at all locations
are identical: $E[Y_i]=\lambda$ ($i=1,...,n$), where $Y_i$ is a random variable
that follows a Poisson distribution with mean $\lambda$:
$(Y_i \mid \lambda) \sim Poisson(\lambda)$.
The observation ($y_i$) is a realization of the random variables $Y$ at site $i$,
these observations are independent and identically distributed (i.i.d.),
and we have $n$ observations in total.

Saying the the distribution is Poisson is an assumption in itself. For example
we assume that the variance equals the mean ($V(\mu)=\mu$).

```{r regr-pois1}
mP0 <- glm(y ~ 1, data=x, family=poisson)
mean(x$y)
mean(fitted(mP0))
exp(coef(mP0))

summary(mP0)
```

The `family=poisson` specification implicitly assumes that we use a logarithmic link functions,
that is to say that $log(\lambda) = \beta_0$, or equivalently: $\lambda = e^{\beta_0}$.
The mean of the observations equal the mean of the fitted values, as expected.

The logarithmic function is called the link function, its inverse, the exponential function
is called the inverse link function. The model family has these convenently stored for us:

```{r regr-family}
mP0$family
mP0$family$linkfun
mP0$family$linkinv
```

## Exploring covariates

Now, in the absence of info about species biology, we are looking at a blank page.
How should we proceed? What kind of covariate (linear predictor) should we use?
We can do a quick and dirty exploration to see what are the likely candidates.
We use a regression tree (`ctree` refers to conditional trees). It is 
a nonparametric method based on binary recursive partitioning in a conditional inference framework.
This means that binary splits are made along the predictor variables,
and the explanatory power of the split is assessed based on how it
maximized difference between the splits and minimized the difference inside the splits.
It is called conditional, because every new split is conditional on the previous splits
(difference can be measured in many different ways, think e.g. sum of squares).
The stopping rule in this implementation is based on permutation tests (see `?ctree` or details
and references).

```{r regr-ctree,fig.width=20,fig.height=15}
mCT <- ctree(y ~ Open + Water + Agr + UrbInd + SoftLin + Roads + 
  Decid + OpenWet + Conif + ConifWet, data=x)
plot(mCT)
```

The model can be seen as a piecewise constant regression, where each bucket (defined by
the splits along the tree) yields a constant predictions based on the mean of the
observations in the bucket. Any new data classified
into the same bucket will get the same value. There is no notion of uncertainty
(confidence or prediction intervals) in this nonparameric model.

But we see something very useful: the proportion of deciduous forest in the landscape
seems to be vary influential for Ovenbird abundance.

## Single covariate

With this new found knowledge, let's fit a parametric (Poisson) linear model
using `Decid` as a predictor:

```{r regr-pois2}
mP1 <- glm(y ~ Decid, data=x, family=poisson)
mean(x$y)
mean(fitted(mP0))
coef(mP1)
```

Same as before, the mean of the observations equal the mean of the fitted values.
But instead of only the intercapt, now we have 2 coefficients estimated.
Our linear predictor thus looks like:
$log(\lambda_i) = \beta_0 + \beta_1 x_{1i}$. This means that expected abundance is 
$e^{\beta_0}$ where `Decid`=0,
$e^{\beta_0}e^{\beta_1}$ where `Decid`=1, 
and $e^{\beta_0+\beta_1 x_{1}}$ in between.

The relationship can be visualized by plotting the fitted values against the predictor,
or using the coefficients to make predictions using our formula:

```{r regr-pois2_plot}
dec <- seq(0, 1, 0.01)
lam <- exp(coef(mP1)[1] + coef(mP1)[2] * dec)
plot(fitted(mP1) ~ Decid, x, pch=19, col="grey")
lines(lam ~ dec, col=2)
rug(x$Decid)
```

The model summary tells us that resudials are not quite right (we would expect
0 median and symmertic tails), in line with residual deviance
being much higher than residual degrees of freedom
(these should be close if the Poisson assumption holds).
But, the `Decid` effect is significant (meaning that the effect size is
large compared to the standard error):

```{r regr-pois2_summary}
summary(mP1)
```

We can compare this model to the null (constant, intercept-only) model:

```{r regr-pois2_aic}
AIC(mP0, mP1)
BIC(mP0, mP1)
model.sel(mP0, mP1)
R2dev(mP0, mP1)
```

AIC uses the negative log likelihood and the number of parameters as penalty.
Smaller value indicate a model that is closer to the (unknowable) true model
(caveat: this statement is true only asymptotically, i.e. it holds for very large
sample sizes). For small samples, we of ten use BIC (more penalty for complex models
when sample size is small), or AICc (as in `MuMIn::model.sel`).

The other little table returned by `R2dev` shows deviance based (quasi) $R^2$ and adjusted
$R^2$ for some GLM classes, just for the sake of completeness. The Chi-squared based
test indicates good fit when the $p$-value is high (probability of being distributed
according the Poisson). 

None of these two models is a particularly good fit in terms
of the parametric distribution.
This, however does not mean these models are not useful for making inferential statements
about ovenbirds. How useful these statements are, that is another question.
Let's dive into cinfidence and prediction intervals a bit.

```{r regr-pois_pred,cache=TRUE}
B <- 2000
alpha <- 0.05

xnew <- data.frame(Decid=seq(0, 1, 0.01))
CI0 <- predict_sim(mP0, xnew, interval="confidence", level=1-alpha, B=B)
PI0 <- predict_sim(mP0, xnew, interval="prediction", level=1-alpha, B=B)
CI1 <- predict_sim(mP1, xnew, interval="confidence", level=1-alpha, B=B)
PI1 <- predict_sim(mP1, xnew, interval="prediction", level=1-alpha, B=B)

## nominal coverage is 95%
sum(x$y %[]% predict_sim(mP0, interval="prediction", level=1-alpha, B=B)[,c("lwr", "upr")]) / nrow(x)
sum(x$y %[]% predict_sim(mP1, interval="prediction", level=1-alpha, B=B)[,c("lwr", "upr")]) / nrow(x)
```

A model is said to have good _coverage_ when the prediction intervals
encompass the right amount of the observations. When the nominal level is 95% ($100 \times (1-\alpha)$,
where $\alpha$ is Type I. error rate),
we expect 95% of the observations fall within the 95% _prediction interval_.
The prediction interval includes the uncertainty around the coefficients
(confidence intervals, uncertainty in $\hat{\lambda}$) and the stochasticity coming from the 
Poisson distribution ($Y_i \sim Poisson(\hat{\lambda})$).

The code above calculate the confidence and prediction intervals for the two models.
We also compared the prediction intervals and the nomial levels, and those were quite
close (ours being a bit more conservative), hinting that maybe the Poisson
distributional assumption is not very bad after all, but we'll come back to this later.

Let's see our confidence and prediction intervals for the two models:

```{r regr-pois_PI}
yj <- jitter(x$y, 0.5)

plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="P0")

polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI0$lwr, rev(PI0$upr)), border=NA, col="#0000ff44")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI0$lwr, rev(CI0$upr)), border=NA, col="#0000ff44")
lines(CI0$fit ~ xnew$Decid, lty=1, col=4)

polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI1$lwr, rev(PI1$upr)), border=NA, col="#ff000044")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI1$lwr, rev(CI1$upr)), border=NA, col="#ff000044")
lines(CI1$fit ~ xnew$Decid, lty=1, col=2)

legend("topleft", bty="n", fill=c("#0000ff44", "#ff000044"), lty=1, col=c(4,2),
  border=NA, c("Null", "Decid"))
```

```{block2, type='rmdexercise'}
**Exercise**

What can we conclude from this plot?

Coverage is comparable, so what is the difference then?
  
Which model should I use for prediction and why? (Hint: look at the non overlapping regions.)
```


## Additive model

Generalized additive models (GAMs) are semiparametric, meaning that
parametric assumptions apply, but responses are modelled more flexibly.

```{r regr-gam}
mGAM <- mgcv::gam(y ~ s(Decid), x, family=poisson)
summary(mGAM)
plot(mGAM)
```

```{r regr-glm_plots}
fitCT <- predict(mCT, x[order(x$Decid),])
fitGAM <- predict(mGAM, xnew, type="response")

plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="P0")
lines(CI0$fit ~ xnew$Decid, lty=1, col=1)
lines(CI1$fit ~ xnew$Decid, lty=1, col=2)
lines(fitCT ~ x$Decid[order(x$Decid)], lty=1, col=3)
lines(fitGAM ~ xnew$Decid, lty=1, col=4)
legend("topleft", bty="n", lty=1, col=1:4,
  legend=c("Null", "Decid", "ctree", "GAM"))
```

```{block2, type='rmdexercise'}
**Exercise**

Play with GAM and other variables to understand response curves:

`plot(mgcv::gam(y ~ s(<variable_name>), data=x, family=poisson))`
```

## Nonlinear terms

We can use polynomial terms to approximate the GAM fit:

```{r regr-pois_poly}
mP12 <- glm(y ~ Decid + I(Decid^2), data=x, family=poisson)
mP13 <- glm(y ~ Decid + I(Decid^2) + I(Decid^3), data=x, family=poisson)
mP14 <- glm(y ~ Decid + I(Decid^2) + I(Decid^3) + I(Decid^4), data=x, family=poisson)
model.sel(mP1, mP12, mP13, mP14, mGAM)
```

Not a surprise that the most complex model won. GAM was more complex than that.


```{r regr-pois_poly_plot}
pr <- cbind(
  predict(mP1, xnew, type="response"),
  predict(mP12, xnew, type="response"),
  predict(mP13, xnew, type="response"),
  predict(mP14, xnew, type="response"),
  fitGAM)
matplot(xnew$Decid, pr, lty=1, type="l",
  xlab="Decid", ylab="E[Y]")
legend("topleft", lty=1, col=1:5, bty="n",
  legend=c("Linear", "Quadratic", "Cubic", "Quartic", "GAM"))
```

Let's see how these affect our prediction intervals:

```{r regr-pois_poly_pi,cache=TRUE}
CI12 <- predict_sim(mP12, xnew, interval="confidence", level=1-alpha, B=B)
PI12 <- predict_sim(mP12, xnew, interval="prediction", level=1-alpha, B=B)
CI13 <- predict_sim(mP13, xnew, interval="confidence", level=1-alpha, B=B)
PI13 <- predict_sim(mP13, xnew, interval="prediction", level=1-alpha, B=B)
CI14 <- predict_sim(mP14, xnew, interval="confidence", level=1-alpha, B=B)
PI14 <- predict_sim(mP14, xnew, interval="prediction", level=1-alpha, B=B)

op <- par(mfrow=c(2,2))
plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="Linear")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI1$lwr, rev(PI1$upr)), border=NA, col="#0000ff44")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI1$lwr, rev(CI1$upr)), border=NA, col="#0000ff88")
lines(CI1$fit ~ xnew$Decid, lty=1, col=4)
lines(fitGAM ~ xnew$Decid, lty=2, col=1)

plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="Quadratic")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI12$lwr, rev(PI12$upr)), border=NA, col="#0000ff44")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI12$lwr, rev(CI12$upr)), border=NA, col="#0000ff88")
lines(CI12$fit ~ xnew$Decid, lty=1, col=4)
lines(fitGAM ~ xnew$Decid, lty=2, col=1)

plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="P0")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI13$lwr, rev(PI13$upr)), border=NA, col="#0000ff44")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI13$lwr, rev(CI13$upr)), border=NA, col="#0000ff88")
lines(CI13$fit ~ xnew$Decid, lty=1, col=4)
lines(fitGAM ~ xnew$Decid, lty=2, col=1)

plot(yj ~ Decid, x, xlab="Decid", ylab="E[Y]",
  ylim=c(0, max(PI1$upr)+1), pch=19, col="#bbbbbb33", main="P0")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(PI14$lwr, rev(PI14$upr)), border=NA, col="#0000ff44")
polygon(c(xnew$Decid, rev(xnew$Decid)),
  c(CI14$lwr, rev(CI14$upr)), border=NA, col="#0000ff88")
lines(CI14$fit ~ xnew$Decid, lty=1, col=4)
lines(fitGAM ~ xnew$Decid, lty=2, col=1)
par(op)
```

## Categorical variables

Categorical variables are expanded into a _model matrix_ before estimation. 
The model matrix usually contains indicator variables for each level
(value 1 when factor value equals a particular label, 0 otherwise)
except for the _reference category_
(check `relevel` if you want to change the reference category).

The estimate for the reference category comes from the intercept,
the rest of the estimates are relative to the reference category.
In the log-linear model example this means a ratio.

```{r regr-pois_cat}
head(model.matrix(~DEC, x))
mP2 <- glm(y ~ DEC, data=x, family=poisson)
summary(mP2)
coef(mP2)
```

The estimate for a non-deciduous landscape is
$e^{\beta_0}$, and it is $e^{\beta_0}e^{\beta_1}$ for deciduous landscapes.
Of course such binary classification at the landscape (1 km$^2$) level
doesn't really makes sense for various reasons:

```{r regr-pois_cat1}
boxplot(Decid ~ DEC, x)
model.sel(mP1, mP2)
R2dev(mP1, mP2)
```

Having estimates for each land cover type improves the model,
but the  model using continuous variable is still better

```{r regr-pois_cat2}
mP3 <- glm(y ~ HAB, data=x, family=poisson)
summary(mP3)

model.sel(mP1, mP2, mP3)
R2dev(mP1, mP2, mP3)
```

The prediction in this case would look like:
$log(\lambda_i)=\beta_0 + \sum_{j=1}^{k-1} \beta_j x_{ji}$, where we have $k$ factor levels
(and $k-1$ indicator variables besides the intercept).

Here is a general way of calculating fitted values or making
predictions based on the design matrix (`X`) and the coefficients (`b`)
(column ordering in `X` must match the elements in `b`)
given a parametric log-linear model `object` and data frame `df`:

```{r regr-pred_general,eval=FALSE}
b <- coef(object)
X <- model.matrix(formula(object), df)
exp(X %*% b)
```

## Multiple main effects

We can keep adding variables to the model in a forwards-selection fashion.
`add1` adds variables one at a time, selecting from the scope defined by the formula:

```{r regr-add1}
scope <- as.formula(paste("~ FOR + WET + AHF +",paste(cn, collapse="+")))
tmp <- add1(mP1, scope)
tmp$AIC_drop <- tmp$AIC-tmp$AIC[1] # current model
tmp[order(tmp$AIC),]
```

It looks like `ConifWet` is the best covariate to add next because it leads to the biggest drop in AIC,
and both effects are significant.

```{r regr-var2}
mP4 <- glm(y ~ Decid + ConifWet, data=x, family=poisson)
summary(mP4)
```

`drop1` is the function opposite of `add1`, it assesses which term should
be dropped from a more saturated model:

```{r regr-var_drop1}
formula_all <- y ~ Open + Agr + UrbInd + SoftLin + Roads + 
  Decid + OpenWet + Conif + ConifWet + 
  OvernightRain + TSSR + DAY + Longitude + Latitude

tmp <- drop1(glm(formula_all, data=x, family=poisson))
tmp$AIC_drop <- tmp$AIC-tmp$AIC[1] # current model
tmp[order(tmp$AIC),]
```

The `step` function can be used to automatically select the best model
based on adding/dropping terms:

```{r regr-var_all,cache=TRUE}
mPstep <- step(glm(formula_all, data=x, family=poisson), 
  trace=0) # use trace=1 to see all the steps
summary(mPstep)
```

## Interaction

When we consider interactions between two variables (say $x_1$ and $x_2$),
we really referring to adding another variable to the model matrix
that is a product of the two variables ($x_{12}=x_1 x_2$):

```{r}
head(model.matrix(~x1 * x2, data.frame(x1=1:4, x2=10:7)))
```

Let's consider interaction between our two predictors from before:

```{r regr-inter}
mP5 <- glm(y ~ Decid * ConifWet, data=x, family=poisson)
summary(mP5)
model.sel(mP0, mP1, mP4, mP5)
```

The model with the interaction is best supported, but how do we make sense of this
relationship? We can't easily visualize it in a single plot. We can either

1. fix all variables (at their mean/meadian) and see how the response is changing along a single variable: this is called a _conditional_ effect (conditional on fixing other variables), this is what `visreg::visreg` does;
2. or plot the fitted values against the predictor variables, this is called a _marginal_ effects, and this is what `ResourceSelection::mep` does.

```{r regr-visreg2}
visreg(mP5, scale="response", xvar="ConifWet", by="Decid")
```
```{r regr-visreg3,fig.show = 'hold',out.width='33%'}
mep(mP5)
```

Let's use GAM to fit a bivariate spline:

```{r regr-GAM2}
mGAM2 <- mgcv::gam(y ~ s(Decid, ConifWet), data=x, family=poisson)
plot(mGAM2, scheme=2, rug=FALSE)
```

Final battle of Poisson models: 

```{r}
model.sel(mP0, mP1, mP12, mP13, mP14, mP2, mP3, mP4, mP5, mGAM, mGAM2)
R2dev(mP0, mP1, mP12, mP13, mP14, mP2, mP3, mP4, mP5, mGAM, mGAM2)
```

Of course, the most complex model wins 
but the Chi-square test is still significant (indicating lack of fit). 
Let's try different error distribution.

## Different error distributions

We will use the 2-variable model with interaction:

```{r regr-dist1}
mP <- glm(y ~ Decid * ConifWet, data=x, family=poisson)
```

Let us try the Negative Binomial distribution first.
This distribution is related to Binomial experiments
(number of trials required to get a fixed number of successes
given a binomial probability). It can also be derived
as a mixture of Poisson and Gamma distributions
(see [Wikipedia](https://en.wikipedia.org/wiki/Negative_binomial_distribution#Gamma%E2%80%93Poisson_mixture)),
which is a kind of hierarchical model.
In this case, the Gamma distribution acts as an i.i.d.
random effect for the intercept:
$Y_i\sim Poisson(\lambda_i)$,
$\lambda_i  \sim Gamma(e^{\beta_0+\beta_1 x_{1i}}, \gamma)$,
where $\gamma$ is the Gamma variance.

The Negative Binomial variance (using the parametrization common in R functions)
is a function of the mean and the scale: $V(\mu) = \mu + \mu^2/\theta$.

```{r regr-dist2}
mNB <- glm.nb(y ~ Decid * ConifWet, data=x)
summary(mNB)
```

Next, we look at zero-inflated models.
In this case, the mixture distribution is a Bernoulli distribution
and a count distribution (Poisson or Negative Binomial, for example).
The 0's can come from both the zero and the count distributions,
whereas the >0 values can only come from the count distribution:
$A_i \sim Bernoulli(\varphi)$, $Y_i \sim Poisson(A_i \lambda_i)$.

The zero part of the zero-inflated models are often parametrized
as probability of zero ($1-\varphi$), as in the `pscl::zeroinfl` function:

```{r regr-dist3}
## Zero-inflated Poisson
mZIP <- zeroinfl(y ~ Decid * ConifWet | 1, x, dist="poisson")
summary(mZIP)

## Zero-inflated Negative Binomial
mZINB <- zeroinfl(y ~ Decid * ConifWet | 1, x, dist="negbin")
summary(mZINB)
```

Now we compare the four different parametric models:

```{r regr-dist4}
AIC(mP, mNB, mZIP, mZINB)
```

Our best model is the Zero-inflated Negative Binomial.
The probability of observing a zero as part of the zero
distribution is back transformed from the zero coefficient
using the inverse logit function:

```{r regr-dist5}
unname(plogis(coef(mZINB, "zero"))) # P of 0
```

Now we use the scale parameter to visualize the variance functions
for the Negative Binomial models (the 1:1 line is the Poisson model):

```{r regr-dist6}
mNB$theta
mZINB$theta

mu <- seq(0, 5, 0.01)
plot(mu, mu + mu^2/mNB$theta, type="l", col=2,
  ylab=expression(V(mu)), xlab=expression(mu))
lines(mu, mu + mu^2/mZINB$theta, type="l", col=4)
abline(0,1, lty=2)
legend("topleft", bty="n", lty=1, col=c(2,4),
  legend=paste(c("NB", "ZINB"), round(c(mNB$theta, mZINB$theta), 2)))
```

```{block2, type='rmdexercise'}
**Exercise**

How can we interpret these different kinds of overdispersion (zero-inflation and higher than Poisson variance)?
  
What are some of the biological mechanisms that can contribute to the 
overdispersion?
```

It is also common practice to consider generalized linear mixed models (GLMMs)
for count data. These mixed models are usually considered as
Poisson-Lognormal mixtures. The simplest, so called i.i.d., case
is similar to the Negative Binomial, but instead of Gamma, we have Lognormal
distribution:
$Y_i\sim Poisson(\lambda_i)$,
$log(\lambda_i) = \beta_0+\beta_1 x_{1i}+\epsilon_i$,
$\epsilon_i \sim Normal(0, \sigma^2)$,
where $\sigma^2$ is the Lognormal variance on the log scale.

We can use the `lme4::glmer` function: use `SiteID` as random effect
(we have exactly $n$ random effects).

```{r regr-dist7,cache=TRUE}
mPLN1 <- glmer(y ~ Decid * ConifWet + (1 | SiteID), data=x, family=poisson)
summary(mPLN1)
```

```{block2, type='rmdnote'}
**Note** 
  
The number of unknowns we have to somehow estimate is now more than the number of observations we have. How is that possible?
```

Alternatively, we can use `SurveyArea` as a grouping variable.
We have now $m < n$ random effects, and survey areas can be seen
as larger landscapes within which the sites are clustered:
$Y_ij\sim Poisson(\lambda_ij)$,
$log(\lambda_ij) = \beta_0+\beta_1 x_{1ij}+\epsilon_i$,
$\epsilon_i \sim Normal(0, \sigma^2)$.
The index $i$ ($i=1,...,m$) defines the cluster (survey area),
the $j$ ($j=1,...,n_i$) defines the sites within survey area $i$
($n = \sum_{i=1}^m n_i$).

```{r regr-dist8,cache=TRUE}
mPLN2 <- glmer(y ~ Decid * ConifWet + (1 | SurveyArea), data=x, family=poisson)
summary(mPLN2)
```

In the battle of distributions (keeping the linear predictor
part the same) the clustered GLMM was best supported:

```{r regr-dist9,cache=TRUE}
tmp <- AIC(mP, mNB, mZIP, mZINB, mPLN1, mPLN2)
tmp$delta_AIC <- tmp$AIC - min(tmp$AIC)
tmp[order(tmp$AIC),]
```


```{block2, type='rmdexercise'}
**Exercise**

What are some of the biological mechanisms that can lead to the 
clustered GLMM bi be the best model?
```


## Count duration effects

Let's change gears a bit now, and steer closer to the main focus
of this book. We want to account for methodological differences
among samples. One aspect of mathodologies involve 
variation in total counting duration. We'll now inspect what
that does to our observations.

First, we create a list of matrices where counts are
tabulated by surveys and time intervals for each species:

```{r regr-time1}
ydur <- Xtab(~ SiteID + Dur + SpeciesID , 
  josm$counts[josm$counts$DetectType1 != "V",])
```

We use the same species (`spp`) as before and create a
data frame indluring the cumulative counts during 3, 5, and 10 minutes:

```{r regr-time2}
y <- as.matrix(ydur[[spp]])
head(y)
colMeans(y) # mean count of new individuals
cumsum(colMeans(y)) # cumulative counts

x <- data.frame(
  josm$surveys, 
  y3=y[,"0-3min"],
  y5=y[,"0-3min"]+y[,"3-5min"],
  y10=rowSums(y))

table(x$y3)
table(x$y5)
table(x$y10)
```

If we fit single-predictor GLMs to these 3 responses, we get
different fitted values, consistent with our mean counts:

```{r regr-time3}
m3 <- glm(y3 ~ Decid, data=x, family=poisson)
m5 <- glm(y5 ~ Decid, data=x, family=poisson)
m10 <- glm(y10 ~ Decid, data=x, family=poisson)
mean(fitted(m3))
mean(fitted(m5))
mean(fitted(m10))
```

Using the multiple time interval data, we can pretend that
we have a mix of methodologies with respect to count duration:

```{r regr-time4}
set.seed(1)
x$meth <- as.factor(sample(c("A", "B", "C"), nrow(x), replace=TRUE))
x$y <- x$y3
x$y[x$meth == "B"] <- x$y5[x$meth == "B"]
x$y[x$meth == "C"] <- x$y10[x$meth == "C"]
boxplot(y ~ meth, x)
sb <- sum_by(x$y, x$meth)
points(1:3, sb[,1]/sb[,2], col=2, type="b", pch=4)
```

We can estimate the effect of the methodology:

```{r regr-time5}
mm <- glm(y ~ meth - 1, data=x, family=poisson)
summary(mm)
exp(coef(mm))
```

Or the effect of the continuous predictor and the method (discrete):

```{r regr-time6}
mm <- glm(y ~ Decid + meth, data=x, family=poisson)
summary(mm)
boxplot(fitted(mm) ~ meth, x)
exp(coef(mm))
```

The fixed effects adjusts the means well:

```{r regr-time7}
cumsum(colMeans(y))
mean(y[,1]) * c(1, exp(coef(mm))[3:4])
```

But it is all relative, depends on reference methodology/protocol.
The problem is, we can't easily extrapolate to a methodology
with count duration of 12 minutes, or interpolate to a mathodology
with count duration of 2 or 8 minutes.
We need somehow to express time expediture in minutes to make that work.
Let's try something else:

```{r regr-time8}
x$tmax <- c(3, 5, 10)[as.integer(x$meth)]
mm <- glm(y ~ Decid + I(log(tmax)), data=x, family=poisson)
summary(mm)
tmax <- seq(0, 20, 0.01)
plot(tmax, exp(log(tmax) * coef(mm)[3]), type="l",
  ylab="Method effect", col=2)
```

Now we are getting somewhere. But still, this function keep
increasing monotonically. 

```{block2, type='rmdexercise'}
**Exercise**

What kind of function would we need and why? 

What is the underlying biological mechanism?
```

## Count radius effects

Before solving the count duration issue, let us look at the 
effect of survey area.
We get a similar count breakdown, but now by distance band:

```{r regr-area1}
ydis <- Xtab(~ SiteID + Dis + SpeciesID , 
  josm$counts[josm$counts$DetectType1 != "V",])

y <- as.matrix(ydis[[spp]])
head(y)
colMeans(y) # mean count of new individuals
cumsum(colMeans(y)) # cumulative counts

x <- data.frame(
  josm$surveys, 
  y50=y[,"0-50m"],
  y100=y[,"0-50m"]+y[,"50-100m"])

table(x$y50)
table(x$y100)
```

We don't consider the unlimited distance case, because the survey area there
is unknown (although we will ultimately address this problem mater).
We compare the counts within the 0-50 and 0-100 m circles:

```{r regr-area2}
m50 <- glm(y50 ~ Decid, data=x, family=poisson)
m100 <- glm(y100 ~ Decid, data=x, family=poisson)
mean(fitted(m50))
mean(fitted(m100))
coef(m50)
coef(m100)
```

## Offsets

Offsets are constant terms in the linear predictor,
e.g. $log(\lambda_i) = \beta_0 + \beta_1 x_{1i} + o_i$,
where $o_i$ is an offset. In the survey area case,
an offset might be the log of area surveyed.
The logic for this is based on point processes:
intensity is a linear function of area
under a homogeneous Poisson point process.
So we can assume that $o_i = log(A_i)$, where $A$ stands for area.

Let's see if using area as offset makes our models comparable:

```{r regr-area3}
m50 <- glm(y50 ~ Decid, data=x, family=poisson, 
  offset=rep(log(0.5^2*pi), nrow(x)))
m100 <- glm(y100 ~ Decid, data=x, family=poisson,
  offset=rep(log(1^2*pi), nrow(x)))
coef(m50)
coef(m100)
mean(exp(model.matrix(m50) %*% coef(m50)))
mean(exp(model.matrix(m100) %*% coef(m100)))
```

These coefficients and mean predictions are much closer to each other,
but something else is going on.

```{block2, type='rmdexercise'}
**Exercise**

Can you guess why we cannot make abundances comparable using
log area as as offset?
```

We pretend again, that survey area varies in our data set:

```{r regr-area4}
set.seed(1)
x$meth <- as.factor(sample(c("A", "B"), nrow(x), replace=TRUE))
x$y <- x$y50
x$y[x$meth == "B"] <- x$y100[x$meth == "B"]
boxplot(y ~ meth, x)
```

Methodology effect:

```{r regr-area5}
mm <- glm(y ~ meth - 1, data=x, family=poisson)
summary(mm)
exp(coef(mm))
```

Predictor and method effects:

```{r regr-area6}
mm <- glm(y ~ Decid + meth, data=x, family=poisson)
summary(mm)
boxplot(fitted(mm) ~ meth, x)
exp(coef(mm))

cumsum(colMeans(y))[1:2]
mean(y[,1]) * c(1, exp(coef(mm))[3])
```

Use log area as continuous predictor:
we would expect a close to 1:1 relationship on the
abundance scale.

```{r regr-area7}
x$logA <- log(ifelse(x$meth == "A", 0.5, 1)^2*pi)
mm <- glm(y ~ Decid + logA, data=x, family=poisson)
summary(mm)
A <- seq(0, 2, 0.01) # in ha
plot(A, exp(log(A) * coef(mm)[3]), type="l",
  ylab="Method effect", col=2)
abline(0, 1, lty=2)
```

The offset forces the relationship to be 1:1
(it is like fixing the `logA` coefficient to be 1):

```{r regr-area8}
mm <- glm(y ~ Decid, data=x, family=poisson, offset=x$logA)
summary(mm)
boxplot(fitted(mm) ~ meth, x)

cumsum(colMeans(y))[1:2]
c(0.5, 1)^2*pi * mean(exp(model.matrix(mm) %*% coef(mm))) # /ha
```

```{block2, type='rmdexercise'}
**Exercise**

Why did we get a `logA` coefficient that was less than 1 when theoretically we should have gotten 1?
```

Predictions using offsets in `glm` can be tricky.
The safest way is to use the matrix product
(`exp(model.matrix(mm) %*% coef(mm) + <offset>)`).
We can often omit the offset, e.g. in the log area case
we can express the prediction per unit area.
If the unit is 1 ha, as in our case, log(1)=0, which means
the mean abundance per unit area can be calculated by
omitting the offsets all together.

