# Dealing with Recordings {#recordings}

## Introduction

Automated recording units (ARU) are increasingly being used for auditory
surveys. There are numerous advantages for using ARUs, e.g.
recordings can be stored in perpetuity to be transcibed 
later, ARUs can be programmed to record at select times and datas over long
time periods that would be prohibitive using human observers.

Bird point counts have beed traditionally done by human observers. Combining
ARU data with traditional point counts thus require an understanding of
how the ARU based counts relate to counts made by human observer in the field.
The best way to approach this question is by simultaneously sampling
by two approaches: (1) human observers doing traditional point count
by registering time and distance interval an individual bird was first detected,
and (2) record the same session at the same location by an ARU to be 
identified/trascribed later in laboratory settings.

## Prerequisites

```{r aru-libs,message=TRUE,warning=FALSE}
library(bSims)                # simulations
library(detect)               # multinomial models
library(mefa4)                # count manipulation
library(paired)               # paired sampling data
```

## Paired data

Human observer surveys:

* 0-50, 50-100, >100 m distance bands,
* 0-3, 3-5, 5-10 minutes time intervals.

ARU surveys:

* unlimited distance,
* 10 minutes survey in 1-minute time intervals.

```{r}
with(paired, ftable(SurveyType, Interval, DISTANCE))
```





We estimate EDR from human observer based counts with distance 
information for species with at least 15 detections making sure that the 
total count for at least some locations exceeded 1:

```{r}
## Data for EDR estimation
xtdis <- Xtab(Count ~ PKEY + DISTANCE + SPECIES, 
  data=paired[paired$SurveyType == "HUM",])
for (i in seq_len(length(xtdis)))
    xtdis[[i]] <- as.matrix(xtdis[[i]][,c("0-49 m", "50-100 m", ">100 m")])

## species subset
nmin <- 15
SPP <- SPP[!(SPP %in%   
  c("CANG","COLO","COGO","COME","FRGU","BCFR","UNKN","RESQ",
  "CORA","AMCR","WOSP","WWCR","PISI","EVGR"))]
xtdis <- xtdis[SPP]
## number of >0 counts
ndis <- sapply(xtdis, function(z) sum(rowSums(z)>0))
## max count
maxd <- sapply(xtdis, function(z) max(rowSums(z)))
xtdis <- xtdis[ndis >= nmin & maxd > 1]

## distance radii
DDdis <- matrix(c(0.5, 1, Inf), nrow(xtdis[[1]]), 3, byrow=TRUE)

## predictors
xdis <- nonDuplicated(paired, PKEY, TRUE)
xdis <- xdis[rownames(xtdis[[1]]),]

## open land cover
xdis$open <- ifelse(xdis$Class_Name %in% c("Open Herb/Grass",
    "Open coniferous","Open Mature Deciduous","Open Mixed",
    "Open Northern","Open Young Deciduous",
    "Open Young Mixed","Poorly Drained"), 1, 0)
```

Preparing objects for removal sampling:

```{r}
```{r,echo=FALSE,comment=NA}
## SRA estimation
xtdurH <- Xtab(Count ~ PKEY + Interval + SPECIES, 
  paired[paired$SurveyType == "HUM",])
xtdurA <- Xtab(Count ~ PKEY + Interval + SPECIES, 
  paired[paired$SurveyType == "ARU",])
DDdur <- matrix(c(3, 5, 10), nrow(xtdurH[[1]]), 3, byrow=TRUE)
xtdurH <- xtdurH[SPP]
xtdurA <- xtdurA[SPP]

xdur <- nonDuplicated(paired, PKEY, TRUE)
xx <- xdur[rownames(xtdurA[[1]]),]

ndurH <- sapply(xtdurH, function(z) sum(rowSums(z)>0))
ndurA <- sapply(xtdurA, function(z) sum(rowSums(z)>0))
maxH <- sapply(xtdurH, function(z) max(rowSums(z)))
maxA <- sapply(xtdurA, function(z) max(rowSums(z)))
OK <- ndurH >= nmin & ndurA >= nmin & maxH > 1 & maxA > 1
xtdurH <- xtdurH[OK]
xtdurA <- xtdurA[OK]
DDdur2 <- rbind(DDdur, DDdur)
```

## The logic

The expected value of the total count (single species) in a 10-minutes time interval
using human observer (subscript $H$) based unlimited radius point count may be written as:
$E[Y_{H}] = D A_{H} p_{H}$ where $Y_{H}$ is the count, $D$ is population density, $A$ is 
the area sampled, $p_{H}$ is the probability that an average individual of the species
is available for detection. The quantity $p_{H}$ can be estimated based on removal 
sampling utilizing the multiple time intervals. $A_{H}$ is often unknown, but can
be estimated using the effective detection radius: $A_{H}=\pi EDR_{H}^2$.
Human observer based EDR is estimated from distance sampling.

The ARU based survey (subscript $R$ for recorder) 
can distinguish individuals within distinct time intervals,
but assigning these individuals is not yet possible using a single ARU.
An ARU based count thus can be seen ans an unlimited radius point count
where the effective area sampled is unknown. The expected value for an
ARU based count for a given species may be written as:
$E[Y_{R}] = D A_{R} p_{R}$. $p_{R}$ can be estimated based on removal 
sampling utilizing the multiple time intervals from the ARU based survey.
The unknown sampling are can be written as $A_{R}=\pi EDR_{R}^2$. 
The problem is that ARU based EDR cannot directly be estimated from the data
because of the lack of multiple distance bands or individial based distance
information.

The advantage of simultaneous sampling by human observers (H) and ARUs (A)
is that population density ($D=D_{H}=D_{R}$) is identical by design.
Possible mechanisms for differences in availability of bird individuals for detection 
($p_{H}$ vs. $p_{R}$) can include differences in how detections
are made in the field vs. in laboratory (e.g. possibility of double checking).

Both $p_{H}$ and $p_{R}$ can be estimated from the data, and the equivalence
$p=p_{H}=p_{R}$ can be tested. So for the sake of simplicity, we assume that
human observer and ARU based $p$'s are equal. 
Dividing the expected values of the counts may be written as:

$$ \frac{E[Y_{R}]}{E[Y_{H}]} = \frac{D A_{R} p}{D A_{R} p} = \frac{\pi EDR_{R}^2}{\pi EDR_{H}^2} = \frac{EDR_{R}^2}{EDR_{H}^2}$$

By substituting $EDR_{R}^2 = \Delta^2 EDR_{H}^2$ (and thus EDR_{R} = \Delta EDR_{H}) we get:

$$ \frac{E[Y_{R}]}{E[Y_{H}]} = \frac{\Delta^2 EDR_{H}^2}{EDR_{H}^2} = \Delta^2 $$

This means that dividing the mean counts from ARU and human observed counts
would give an estimate of the squared scaling constant ($\Delta^2$) describing the
relationship between the estimated $EDR_{H}$ and the unknown $EDR_{R}$.





## Estimating the scaling constant

Counts are often modelled in a log-linear Poisson GLM. We used GLM to estimate
the unknow scaling constant from simultaneous (paired) surveys. The poisson mean
for a count made at site $i$ by human observer is 
$\lambda_{i,H} = D_{i} \pi EDR_H^2 p$. $EDR_H$ and $p$ are estimated using distance 
sampling and removal sampling, respectively. Those estimates are used to
calculate a correction factor $C = \pi EDR_H^2 p$ which is used as an offset
on the log scale as $log(\lambda_{i,H}) = log(D_{i}) + log(C) = \beta_0 + log(C)$,
where $\beta_0$ is the intercept in the GLM model.

Following the arguments above, the poisson mean for an ARU based count 
made at site $i$ is $\lambda_{i,R} = D_{i} \pi \Delta^2 EDR_H^2 p = D_{i} \delta^2 C$.
On the log scale, this becomes
$log(\lambda_{i,R}) = log(D_{i}) + log(\Delta^2) + log(C) = \beta_0 + \beta_1 + log(C)$,
where $\beta_1$ is a contrast for ARU type surveys in the log-linear model.

We used survey type as a binary variable ($x_i$) with value 0 for human observers
and value 1 for ARUs. So the Poisson model is generalized as:
$log(\lambda_{i}) = \beta_0 + x_i \beta_1 + log(C)$. $\Delta$ can be 
calculated from $\beta_1$ as $\Delta = \sqrt{exp(\beta_i)}$.

## Distance sampling

```{r}
mdis <- list()
for (i in 1:length(xtdis)) {
    m <- cmulti(xtdis[[i]] | DDdis ~ 1, xdis, type = "dis")
    if (!inherits(m, "try-error"))
        mdis[[names(xtdis)[i]]] <- m
}
mdis2 <- list()
for (i in 1:length(xtdis)) {
    m0 <- cmulti(xtdis[[i]] | DDdis ~ 1, xdis, type = "dis")
    m1 <- cmulti(xtdis[[i]] | DDdis ~ open, xdis, type = "dis")
    mdis2[[i]] <- list(m0=m0, m1=m1)
}

t(sapply(mdis2, function(z) coef(z$m1)))

#mdis <- lapply(xtdis, function(z) cmulti(z | DDdis ~ 1, type = "dis"))
edrH <- sapply(mdis, function(z) exp(coef(z)))
names(edrH) <- names(xtdis)
edrH <- edrH[!(names(edrH) %in% c("RESQ", "UNKN"))]
tax <- read.csv("taxonomy-bam.csv")
rownames(tax) <- tax$Species_ID
```

```{r,echo=FALSE,comment=NA}
data.frame(Species=tax[names(edrH),"English_Name"], EDR_H_meters=round(100*edrH, 1))
```

#### Estimating availability

We estimated availability for human observer and ARU based counts 
using the time interval information. ARU based intervals were
collaped to the 0-3-5-10 minutes intervals to match the human observer based
design. We estimated availability for species with at least 15 detections
in both subsets of the data (making sure that the total count for at least
some locations exceeded 1). We analyzed the human observer and ARU based data 
in a single model using survey type as a dummy variable. 
We tested if the estimate corresponding to survey type differed significantly
from 0 using 95\% confidence intervals.

The following table lists singing rates (`phi` 1/minute), probability of
singing in a 10-minutes interval (`p10`), number of detections (`n`),
and whether or not the confidence limits for the survey type estimate
($\beta_1$) contained 0 (i.e. not significant survey effect).

```{r}
mdurA <- list()
mdurH <- list()
mdurHA <- list()
mdurHA1 <- list()
for (i in 1:length(xtdurA)) {
    yA <- as.matrix(xtdurA[[i]])[,c("0-3 min","3-5 min","5-10 min")]
    yH <- as.matrix(xtdurH[[i]])[,c("0-3 min","3-5 min","5-10 min")]
    mdurA[[names(xtdurA)[i]]] <- cmulti(yA | DDdur ~ 1, type = "rem")
    mdurH[[names(xtdurA)[i]]] <- cmulti(yH | DDdur ~ 1, type = "rem")
    yyy <- rbind(yH, yA)
    aru01 <- rep(0:1, each=nrow(xtdurH[[i]]))
    mdurHA[[names(xtdurA)[i]]] <- cmulti(yyy | DDdur2 ~ 1, type = "rem")
    mdurHA1[[names(xtdurA)[i]]] <- cmulti(yyy | DDdur2 ~ aru01, type = "rem")
}
cfA <- sapply(mdurA, coef)
names(cfA) <- names(mdurA)
cfH <- sapply(mdurH, coef)
names(cfH) <- names(mdurH)
cfHA <- sapply(mdurHA, coef)
names(cfHA) <- names(mdurHA)
phiA <- exp(cfA)
phiH <- exp(cfH)
phiHA <- exp(cfHA)

cfHA1 <- t(sapply(mdurHA1, coef))

ci <- t(sapply(mdurHA1, function(z) confint(z)[2,]))
nam <- rownames(ci)[rowSums(is.na(ci)) > 0 | rowSums(sign(ci)) != 0]

#plot(phiA ~ phiH, ylim=c(0,0.6), xlim=c(0,0.6))
#abline(0,1)
#abline(lm(phiA ~ phiH - 1), col=2)
#abline(lm(phiA ~ phiH), col=4)
#confint(lm(phiA ~ phiH - 1))
#confint(lm(phiA ~ phiH))

data.frame(Species=tax[rownames(cfHA1),"English_Name"], 
    phi_H=round(exp(cfHA1[,1]), 3),
    phi_A=round(exp(cfHA1[,1]+cfHA1[,2]), 3),
    p10_H=round(1-exp(-10*exp(cfHA1[,1])), 3),
    p10_A=round(1-exp(-10*exp(cfHA1[,1]+cfHA1[,2])), 3),
    p_equal=rowSums(is.na(ci)) > 0 | rowSums(sign(ci)) == 0,
    n_A=ndurA[rownames(ci)], 
    n_H=ndurH[rownames(ci)])
```

```{r}
mdur <- list()
for (i in 1:length(xtdurA)) {
    if (interactive())
        cat(SPP[i], "\n")
    yA <- as.matrix(xtdurA[[i]])[,c("0-3 min","3-5 min","5-10 min")]
    yH <- as.matrix(xtdurH[[i]])[,c("0-3 min","3-5 min","5-10 min")]
    yyy <- rbind(yH, yA)
    aru01 <- rep(0:1, each=nrow(xtdurH[[i]]))
    Noise <- c(xx$Noise, xx$Noise)
    mdur[[i]] <- list()
    m0 <- cmulti(yyy | DDdur2 ~ 1, type = "rem")
    m1 <- cmulti(yyy | DDdur2 ~ aru01, type = "rem")
    m2 <- cmulti(yyy | DDdur2 ~ Noise, type = "rem")
    m3 <- cmulti(yyy | DDdur2 ~ aru01 + Noise, type = "rem")
    m4 <- cmulti(yyy | DDdur2 ~ aru01 * Noise, type = "rem")
    aic <- AIC(m0, m1, m2, m3, m4)
    aic$dAIC <- aic$AIC - min(aic$AIC)
    mdur[[i]] <- list(m0=m0, m1=m1, m2=m2, m3=m3, m4=m4, aic=aic)
}

names(mdur) <- names(xtdurA)
table(best <- sapply(mdur, function(z) which.min(z$aic$dAIC)-1))
names(best) <- names(xtdurA)

t(sapply(mdur, function(z) coef(z$m1)))
t(sapply(mdur[best==2], function(z) coef(z$m2)))

```

#### Estimating survey type effect

We focused on species with reasonable sample sizes ($n > 14$), and where
we did not find significant survey type effect on availability.
We used the Poisson GLM model describe before to estimate the $\beta_1$
coefficient corresponding to survey type as binary predictor variable,
and an offset term incorporating human observer based effective area sampled and 
availability.

```{r,echo=FALSE}
## these are the species where we have:
## - reasonable sample size from training data
##   to estimate phi and edr
## - phi is not significantly different between aru and human
namOK <- rownames(ci)[rowSums(sign(ci)) == 0]
SPP2 <- intersect(names(edrH), namOK)
phiOK <- phiHA[SPP2]
edrOK <- edrH[SPP2]
round(phiOK,4)
round(edrOK,4)

xt <- as.matrix(Xtab(Count ~ PKEYm + SPECIES, x))
xxx <- nonDuplicated(x, PKEYm, TRUE)
xxx <- xxx[rownames(xt),]
xxx$distur <- ifelse(xxx$Disturbance != "Undisturbed", 1, 0)

library(lme4)
mods <- list()
cid <- matrix(NA, length(SPP2), 2)
rownames(cid) <- SPP2
colnames(cid) <- c("lcl", "ucl")
aictab <- list()
#spp <- "ALFL"
for (spp in SPP2) {
    y <- xt[,spp]
    C <- edrOK[spp]^2 * pi * (1-exp(-phiOK[spp]))
    off <- rep(log(C), nrow(xxx))
    mod0 <- glm(y ~ 1, xxx, offset=off, family=poisson)
    mod1 <- glm(y ~ SurveyType, xxx, offset=off, family=poisson)
    mod2 <- glm.nb(y ~ SurveyType + offset(off), xxx)
    mod3 <- glmer(y ~ SurveyType + (1|PKEY), xxx,
        offset=off, family=poisson)
    mod4 <- glmer(y ~ SurveyType + distur + (1|PKEY), xxx,
        offset=off, family=poisson)
    mod5 <- glmer(y ~ SurveyType * distur + (1|PKEY), xxx,
        offset=off, family=poisson)
    
    #summary(mod)
    mods[[spp]] <- mod3
    bb <- mvrnorm(10^4, fixef(mod3), vcov(mod3))
    cid[spp,] <- quantile(sqrt(exp(bb[,"SurveyTypeARU"])), c(0.025, 0.975))
    aic <- AIC(mod0, mod1, mod2, mod3, mod4, mod5)
    aic$dAIC <- aic$AIC - min(aic$AIC)
    aictab[[spp]] <- aic
}

aic_support <- t(sapply(aictab, function(z) z[,3]))

delta <- sqrt(exp(sapply(mods, function(z) fixef(z)[2])))
names(delta) <- names(mods)
```

The following table show the estimate of $\delta$ for each species,
and the corresponding estimates of effective detection radius (EDR) in meters
and effective area sampled (A) in ha:

```{r,echo=FALSE,comment=NA}
data.frame(Species=tax[SPP2,"English_Name"], 
    round(cbind(EDR_H=100*edrOK, EDR_A=100*delta*edrOK, 
        A_H=edrOK^2*pi, A_A=(delta*edrOK)^2*pi,
        delta=delta, delta_sq=delta^2), 2))
```


## Habitat effects

We explored if habitat effects are important to consider for density (as main effect)
and for the survey type effect (as an interaction between habitat ans survey type).
We used canopy closure as continuous predictor describing an important feature of the
habitats that can affect burd density and also sound attenuation with distance.
We compared the following models based on AIC:

1. survey type as only main effect,
2. survey type main effect \& canopy closure main effect,
3. survey type main effect \& canopy closure main effect, \& interaction between the two,

We calculated delta-AIC values for the 3 models for the species.
We found that the interaction (i.e. different EDR ration in different habitats)
was the best supported model only for Common Raven.

```{r,echo=FALSE,comment=NA}
## explore habitat effects
## Codes for veg are e.g. tA= trembling aspen. 
## wS= white spruce. Most should be self-explanatory, 
## exceptions are wI = willow aL = alder
Aic <- list()
for (spp in SPPv) {
    i <- !is.na(xxx$Canopy.Closure)
    y <- xt[i,spp]
    C <- edrOK[spp]^2 * pi * (1-exp(-phiOK[spp]))
    off <- rep(log(C), length(y))
    m0 <- glm(y ~ SurveyType, xxx[i,], offset=off, family=poisson)
    m1 <- update(m0, . ~ . + Canopy.Closure)
    m2 <- update(m0, . ~ . * Canopy.Closure)
    aic <- AIC(m0, m1, m2)
    aic$dAIC <- aic$AIC - min(aic$AIC)
    #aic[order(aic$dAIC),]
    Aic[[spp]] <- aic
}
daic <- t(sapply(Aic, "[[", "dAIC"))
colnames(daic) <- c("model_1","model_2","model_3")
data.frame(Species=tax[SPPv,"English_Name"],
    round(daic, 3))
```


```{r,eval=FALSE,echo=FALSE}
## explore species traits
tr <- read.delim("BirdFuncDat.txt") 
tr <- tr[!duplicated(tr$Scientific),]
rownames(tr) <- tolower(as.character(tr$Scientific))
tax <- read.csv("taxonomy-bam.csv")
rownames(tax) <- tax$Species_ID
tax <- tax[SPP,c("English_Name","Scientific_Name")]
sn <- tolower(as.character(tax[,2]))
sn <- gsub("setophaga", "dendroica", sn)
sn <- gsub("geothlypis", "oporornis", sn)
sn[sn == "troglodytes hiemalis"] <- "troglodytes troglodytes"
tr <- tr[sn,]
tax$mass <- tr$BodyMass.Value
tax$log_mass <- log(tax$mass)
setdiff(sn, rownames(tr))
tax$d <- d[SPP]
tax$log_d <- log(tax$d)
plot(log_d ~ log_mass, tax)
abline(lm(log_d ~ log_mass, tax))
summary(lm(log_d ~ log_mass, tax))
summary(lm(edrOK[SPP] ~ log_mass, tax))
plot(edrOK[SPP] ~ log_mass, tax)
abline(lm(edrOK[SPP] ~ log_mass, tax))
```

## Next steps

* check availability models and make sure that estimates make sense
  (otherwise, use location specific models -- and corresponding
  offsets in final modeling)
* EDR models: see of LCC/disturbance based models are needed,
  make sense etc -- use corresponding offsets
* final model needs to recognize location/survey specific offsets
* final model Poisson-Ln mixed model with random intercept for pairs,
  plus include some LCC/disturbance based fixed effects
* if SurveyType * habitat interactions are concerned: keep
  a simplified model as well
* Noise level: make sure model and validation data are balanced
  (we can then check if noise is driving some of the outliers etc)
  Site & noise level: might be OK for now, let's revisit when
  doing the validation.


## Others


fixed effects: using paired design

fixed effects: pretend that design is not paired
- random sampling
- biased sampling by habitat





