---
title: "R Basics"
date: "Point count data analysis workshop, AOS 2019, Anchorage AK, 24 June 2019"
author: "Peter Solymos"
output: pdf_document
fontsize: 11pt
classoption: letterpaper
urlcolor: blue
---

This short document is intended to help you brush up your R skills.
If you feel that these R basics are not very familiar,
I suggest to take a look at some introductory R books,
sich as this preprint version of Norman Matloff's *The Art of R Programming* book: http://heather.cs.ucdavis.edu/~matloff/132/NSPpart.pdf, check out Chapters 1--6.


## R is a great calculator

```{r}
1 + 2
```



## Assign a value and print an object

```{r}
(x = 2) # shorthand for print
print(x)
x == 2 # logical operator, not assignment
y <- x + 0.5
y # another way to print
```



## Logical operators

```{r}
x == y # equal
x != y # not eaqual
x < y # smaller than
x >= y # greater than or equal
```



## Vectors and sequences

```{r}
x <- c(1, 2, 3)
x
1:3
seq(1, 3, by = 1)

rep(1, 5)
rep(1:2, 5)
rep(1:2, each = 5)
```



## Vector operations, recycling

```{r}
x + 0.5
x * c(10, 11, 12, 13)
```



## Indexing vectors, ordering

```{r}
x[1]
x[c(1, 1, 1)] # a way of repeatig values
x[1:2]
x[x != 2]
x[x == 2]
x[x > 1 & x < 3]
order(x, decreasing=TRUE)
x[order(x, decreasing=TRUE)]
rev(x) # reverse
```



## Character vectors, `NA` values, and sorting

```{r}
z <- c("b", "a", "c", NA)
z[z == "a"]
z[!is.na(z) & z == "a"]
z[is.na(z) | z == "a"]
is.na(z)
which(is.na(z))
sort(z)
sort(z, na.last=TRUE)
```



## Special values

```{r}
as.numeric(c("1", "a")) # NA: not available (missing or invalid)
0/0 # NaN: not a number
1/0 # Inf
-1/0 # -Inf
```


## Matrices and arrays

```{r}
(m <- matrix(1:12, 4, 3))
matrix(1:12, 4, 3, byrow=TRUE)

array(1:12, c(2, 2, 3))
```



## Attribues

```{r}
dim(m)
dim(m) <- NULL
m
dim(m) <- c(4, 3)
m
dimnames(m) <- list(letters[1:4], LETTERS[1:3])
m
attributes(m)
```



## Matrix indices

```{r}
m[1:2,]
m[1,2]
m[,2]
m[,2,drop=FALSE]
m[2]

m[rownames(m) == "c",]
m[rownames(m) != "c",]
m[rownames(m) %in% c("a", "c", "e"),]
m[!(rownames(m) %in% c("a", "c", "e")),]
```



## Lists and indexing

```{r}
l <- list(m = m, x = x, z = z)
l
l$ddd <- sqrt(l$x)
l[2:3]
l[["ddd"]]
```



## Data frames

```{r}
d <- data.frame(x = x, sqrt_x = sqrt(x))
d
```



## Structure

```{r}
str(x)
str(z)
str(m)
str(l)
str(d)
str(as.data.frame(m))
str(as.list(d))
```



## Summaries

```{r}
summary(x)
summary(z)
summary(m)
summary(l)
summary(d)
```



## Key takehome

A *matrix* is a vector with `dim` attribute, elements are in same mode. 

A *data frame* is a list where length of elements match and elements can be in different mode.

