---
title: "Estimating effective detection radius from playback"
author: "Solymos, Leston, Yip, & Bayne"
date: "August 28, 2015"
output: pdf_document
---

## Distance function

The shape of the distance function describes how detection probability
attenuates as a function of distance from observer ($d$).
The distance function is a strictly monotinic decreasing function with
increasing distance. There are many different mathematical formulations
to describe the shape. We choose the half-Normal distance function
because of its simplicity, and because its standard deviation parameter ($\tau$)
is directly interpretable as effective detection radius (EDR)
for unlimited (i.e. not truncated) point counts in bird surveys.
The distance function can be written as: $p(d) = exp(-d^2 / \tau^2)$.

```{r,echo=FALSE,fig.cap='Distance function'}
d <- 0:200
plot(d, exp(-d^2/60^2), type="l", col=1, lwd=2,
    xlab="Distance (m)", ylab="Detection probability, p(d)")
lines(d, exp(-d^2/80^2), type="l", col=2, lwd=2)
lines(d, exp(-d^2/100^2), type="l", col=4, lwd=2)
lines(c(60,60), c(0, exp(-1)), lty=2, col=1)
lines(c(80,80), c(0, exp(-1)), lty=2, col=2)
lines(c(100,100), c(0, exp(-1)), lty=2, col=4)
legend("topright", bty="n", lty=1, col=c(1,2,4), lwd=2,
    legend=c("EDR = 60 m", "EDR = 80 m", "EDR = 100 m"))
```

EDR is an important distance, because for unlimited counts, the area of the point
count is unknown. EDR gives the radius of the circle where the expected value
of available individuals not detected equals the expected value of the
detected individuals outside of that distance. Thus, the effective area
sampled can be estimated from EDR as $\hat{A} = \hat{\tau}^2 \pi = \hat{EDR}^2 \pi$.

## Playback data and model

Playback based data consists of souds transmitted from various known distances.
This then recorded and processed in the lab. Simultaneously, a human observer
is making detections in the field.
The results is a binary reponse variable ($Y \in (0, 1)$). 
The distribution this random variable can be discribed by a Bernoulli 
distribution: $(Y \mid d) \sim Bernoulli(p(d))$, where $p(d)$ follows the 
distance function with the unknown parameter $\tau$.

## Estimating EDR from playback data

Bernoulli type data isu usually analyzed using logistic regression, or
Binomial GLM (with Binomial size = 1) using a logistic link function.
This approach, although totally acceptable, would not lead to an
instant estimate of EDR. Estimating EDR would involve numerical integration
using the estimated logistic based distance function to find the radius
where the volume above the curve inside of EDR equals the volume under the curve
outside of EDR.

Instead of Binomial GLM followed by numerical integration, we choose to
use a different link function.

### Log link

By taking the log of the distance function we get: 
$log(p(d)) = -d^2 / \tau^2 = x \frac{1}{\tau^2} = 0 + x \beta$.
This means that by (1) using a log-link, and (2) fixing the intercept at 0,
we can estimate a function of EDR using a simple linear modelling framework.
The linear predictor is a transformation of the known distance: $x = -d^2$.
The estimate $\hat{\beta}$ can be used to calculate EDR as:
$\hat{\tau}=\sqrt{1/\hat{\beta}}$.

### Complementary log-log link

A complementary log-log link is often used to analyze 0/1 data interpreted as
a right-censored Poisson distribution.

This result highlight that the complementary log-log link (cloglog) can 
be used to approximate the log-link in logistic regression.

Factors and covariates affecting EDR can be incorporated as interactions 
*without* main effects in the model.

## Simulation study, proof of concept

First we simulat data under the known model using EDR = 80 m.

```{r}
n <- 500
tau <- 80 # edr in meters
d <- sample(0:400, n, replace=TRUE) # distance in meters
x <- -d^2 # transformed distance
p <- exp(-d^2 / tau^2) # probs for half-normal distance function
y <- rbinom(n, 1, p)
```

The simulated data is then analyzed using the `glm` function in R as follows:

```{r}
m <- glm(y ~ x - 1, family=binomial("cloglog"))
(edr <- sqrt(1/coef(m)))
```

We can compare the true and estimated distance function to see
how well those agree (on the probability and on the log scale).


```{r,echo=FALSE}
op <- par(mfrow=c(1,2))
plot(d[order(d)], p[order(d)], type="l", lwd=2,
     xlab="Distance (m)", ylab="Probability")
rug(d[y==0], side=1)
rug(d[y==1], side=3)
lines(d[order(d)], exp(-d[order(d)]^2 / edr^2), col=2, lwd=2)

plot(d[order(d)]^2, log(p[order(d)]), type="l", lwd=2,
    xlab="-Distance^2", ylab="log(Probability)")
rug(d[y==0]^2, side=1)
rug(d[y==1], side=3)
lines(d[order(d)]^2, log(exp(-d[order(d)]^2 / edr^2)), col=2, lwd=2)
legend("topright", bty="n", lty=1, col=c(1,2), lwd=2,
    legend=c("Truth", "Estimate"))
par(op)
```

## Parametric bootstrap to compare habitat specific ERDs

## Simulation study, proof of concept

Simulat data under the known model using EDR = 80 m in Habitat A and EDR = 100 m in Habitat B.

```{r}
n <- 500
tauA <- 50
tauB <- 150
d <- sample(0:400, n, replace=TRUE) # distance in meters
x <- -d^2 # transformed distance
H <- as.factor(rep(c("A","B"), each=n/2))
p <- exp(-d^2 / ifelse(H=="A", tauA, tauB)^2)
y <- rbinom(n, 1, p)
```

The simulated data is then analyzed using the `glm` function in R as follows:

```{r}
df <- data.frame(y, H01=ifelse(H=="B", 1, 0), x)
m <- glm(y ~ x + x:H01 - 1, data=df, family=binomial("cloglog"))
summary(m)
cf <- coef(m)
cf[-1] <- cf[-1]+cf[1]

(edr <- sqrt(1/cf))
```

Simulate from the fitted model:
```{r warning=FALSE,message=FALSE}
f <- fitted(m)

bres <- list()
for (i in 1:100) {
    y_star <- rbinom(length(f), 1, f)
    df_star <- data.frame(y=y_star, H01=ifelse(H=="B", 1, 0), x)
    m_star <- glm(y ~ x + x:H01 - 1, data=df_star, family=binomial("cloglog"))
    cf_star <- coef(m_star)
    cf_star[-1] <- cf_star[-1]+cf_star[1]
    
    (edr_star <- sqrt(1/cf_star))
    bres[[i]] <- edr_star
}

bres_mat <- do.call(rbind, bres)
apply(bres_mat, 2, quantile, c(0.05, 0.95))
```

